\subsection{Live Programming for Debugging}

Previous examples highlighted how running different kinds of incomplete
programs---ones with missing expressions, type-inconsistent expressions, or
missing types---can be useful.
%
Our final example demonstrates how expression holes may be useful even for
complete programs, in order to facilitate debugging and program understanding.

%% \overviewExample{6}{\rkc{...}}
%%
%% \rkc{breakpoints, understanding input/output behavior}

\overviewExample{6}{Quicksort}

Consider the following buggy implementation of quicksort, which contains several
errors.
%
The student wants to debug the implementation and so inserts an expression
around the return expression on line \rkc{XXX}, and runs
\li{quicksort [X,X,X,X,X,X]}.

\begin{lstlisting}
quicksort : list('a) -> list('a)
quicksort [] = []
quicksort (x:xs) =
  let (left, right) = span ((>) x) xs in
  let (left', right') = (quicksort left, quicksort right) in
  ? (left' @ [x] @ right) ?
\end{lstlisting}

\noindent
%
Based on the live environment for the initial call, the student notices that the
values in \li{left} are larger than the pivot \li{x}, but \li{left} was intended
to bind smaller values.
%
This suggests that the student got the filtering predicate backwards, so he
replaces it with \li{(<)} on line \rkc{XXX}.
%
After re-running the program and viewing the live output again, \li{left} now
contains smaller values than \li{x}, but not all of them; \li{right} has some
values that are larger and some that are smaller.
%
The student uses type-based search to find other functions, besides \li{span},
of type \li{('a -> bool) -> 'a list -> ('a list, 'a list)} and finds
\li{partition}.
%
It may be easier for the student to simply try the function rather than studying
the documentation, so he edits line \rkc{XXX} to call \li{partition} instead,
re-runs, and observes that both \li{left} and \li{right} now seem to achieve the
intended partitioning.

Despite these two fixes, the result expression is not yet sorted.
%
In particular, the values greater than the pivot \li{x} are not completely
sorted.
%
Viewing the values of bindings of \li{left'} and \li{right'}, the student sees
that both seem to have correct prefixes of sorted lists, but that they are
incorrect after the pivots.
%
This suggests that something may be wrong with the handling of \li{right'}.
%
Indeed, the student takes a look for uses of \li{right'} and sees that line
\rkc{XXX} mistakenly uses \li{right} instead.
%
Making this change and re-running now produces the desired result, built up as a
hole environment tree that shows how all of the recursive results will be
combined by a series of nested appends.
%
Happy with the results of the program, the student removes the hole around the
expression on line \rkc{XXX}, and evaluation can proceed to perform the nested
appends, producing the final sorted result.
%
\autoref{sec:discussion} discusses how debugging tools may be designed to
systematically insert hole expressions in particular places to achieve certain
debugging and program understanding strategies.
%
%% When run on a larger input, this projection of the full evaluation trace shows a
%% nesting structure---which could be rendered as visualization to help
%% understanding, as described in \autoref{sec:discussion}---whose depth is clearly
%% much smaller than the length of the input list; this suggests that the
%% algorithm, indeed, achieves some sort of sublinear complexity, as intended.
%

\begin{lstlisting}
quicksort : list('a) -> list('a)
quicksort [] = []
quicksort (x:xs) =
  let (left, right) = partition ((<) x) xs in
  let (left', right') = (quicksort left, quicksort right) in
  ? (left' @ [x] @ right') ?
\end{lstlisting}
